Effects of COVID-19 related social media use on well-being

### 2023-07-11 manuscript.tex
Major updates Implemented feedback

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={COVID-19 related social media use slightly decreases well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{COVID-19 related social media use slightly decreases well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\shorttitle{COVID-19 related social media use \& well-being}

@@

\abstract{%
In times of crisis such as the COVID-19 pandemic, citizens need to stay informed about recent political events. To this end, people increasingly use social media. However, being particularly engaging, many find it hard to disconnect from social media, especially during times of crisis. Using data from the Austrian Corona Panel Project consisting of 3,485 participants from 32 waves, controlling for several stable and varying confounders, the results showed that COVID-19 related social media use reduced facets of well-being. However, the effects were very small. Other factors such as health, income, exercise, or internal locus of control showed larger and more meaningful effects.
}

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus?
In what region is it spreading?
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, which were at an all time high (Statista, 2021).

Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged.
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children (Klein, 2021).
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health (Sandstrom et al., 2021).
Although research has begun addressing this question (Bendau et al., 2021; Bradley \& Howard, 2021; Choi \& Choung, 2021; Dörnemann et al., 2021; Eden et al., 2020; Guazzini et al., 2022; Latikka et al., 2022; Liu \& Tong, 2020; Riehm et al., 2020; Sewall et al., 2021; Stainback et al., 2020; Yue et al., 2022), it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
By analyzing large-scale panel data from Austria and by focusing on intra-individual causal effects, this study aims to answer to this question.

@@

This study investigates how different \emph{facets} of well-being are affected by different \emph{types} and different \emph{channels} of communication (Meier \& Reinecke, 2020).
Building on the typology of subjective well-being (Diener et al., 2018), three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used (Verduyn et al., 2015), I further distinguish three types of use and five popular channels.
The types of use include reading, posting, and liking and sharing COVID-19 related posts.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

How could the various types and channels of COVID-19 related social media use affect well-being?
According to the set-point theory, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time routinely returning to prior levels (Sheldon \& Lucas, 2014).
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being (Lucas, 2007).
Can media use be such a factor?
Current literature overviews suggest no:
Although social media use on average decreases well-being, the effects are small (Meier \& Reinecke, 2020).
In addition, the effects of media use differ across individuals and types of content (Valkenburg \& Peter, 2013).
Whereas for some users social media are more beneficial, for others they are more harmful.
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral (Beyens et al., 2021).
This finding is aligned with the Differential Susceptibility to Media Effects Model:
Although there is substantial \emph{variation} of media effects for individual users, there are both negative and positive aspects, leading to small effects on average (Valkenburg \& Peter, 2013)---as confirmed by several literature overviews (Dienlin \& Johannes, 2020; Huang, 2017; Meier \& Reinecke, 2020; Orben, 2020).

Why are the effects of social media use on well-being small on average?
Using Social media is ambivalent.
It can impair well-being when causing embarrassment, stress, or disinformation, and it can improve well-being when providing connectedness, information, or entertainment (Büchi, 2021).
Although social media are often associated with negative outcomes, there are a large number of positive effects, which either offset or at least significantly reduce negative outcomes.
Positive outcomes encompass finding relevant information; maintaining and fostering relationships; expressing one's personality; or entertaining oneself (e.g., Pelletier et al., 2020).
Negative effects include distraction; displacement of other meaningful activities; consumption of shallow, misleading or challenging content; or negative social comparisons (e.g., Meier \& Krause, 2022).
People are much more likely to use media that offer plentiful benefits (Katz et al., 1973).
Users implicitly learn what media help them regulate their mood and thereby well-being (Zillmann, 1988).
Given the ubiquitous use of social media, the many benefits they offer, plus the general and implicit competence to use media that foster mood management, it is hence unlikely that the effects of social media use on well-being are profusely negative.

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Results are mixed.
When browsing social media for COVID-19 related news, many users reported being captivated to such an extent they could not stop using social media (Klein, 2021).
In a study with 1,131 residents from Wuhan in China, people who spent more time in quarantine also spent more time on social media (Yue et al., 2022).
Those who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.
People who used social media as a primary source of information reported on average ``significantly more unspecific anxiety and depression {[}\ldots{]} and significantly more specific COVID-19 related anxiety symptoms'' (Bendau et al., 2021, p. 288).
Eden et al. (2020) analyzed how 425 US college students used media during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a large-scale study with 11,537 respondents from the US, increased COVID-19-related media consumption was related to more psychological distress (Stainback et al., 2020).
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use during the pandemic on mental health, finding that digital technology did not have significant effects on mental health (Sewall et al., 2021; for a similar study with comparable results, see Bradley \& Howard, 2021).
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness increased (Guazzini et al., 2022).
In a study with 735 participants from Finland, levels of loneliness were stable during the pandemic, and people who engaged more on social media experienced less loneliness (Latikka et al., 2022).
Together, the literature is mixed, showing both positive and negative affects of social media use focused on or during COVID-19 on well-being (see also Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020).

In conclusion, these mixed empirical results, together with the observation that social media effects on well-being are very small in general, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

\begin{quote}
Hypothesis: The within-person effects of all types of COVID-19 related social media use on all types of well-being indicators---while controlling for several stable and varying covariates such as sociodemographic variables and psychological dispositions---will be trivial.
\end{quote}

@@

The data come from the Austrian Corona Panel Project (Kittel et al., 2021), which is a large-scale standalone panel study.
The data are hosted on AUSSDA, are publicly available here (\url{https://doi.org/10.11587/28KQNS}), and consist of 32 waves.
The study was conducted between March 2020 and June 2022, and data collection is now finished.
Between March 2020 and July 2020, the intervals between waves were weekly and afterward monthly.
Each wave consists of at least 1,500 respondents.
The sample size was \emph{N} = 3,485, with overall 111,520 observations.
Panel mortality was compensated through a continuous re-acquisition of new participants.
Participants were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Panel members were incentivized with 180 credit points for each wave of the study.

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age.
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets).
Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements.
The participants provided their written informed consent to participate in this study.
The average age was 41 years, 49 percent were male, 14 percent had a University degree, and 5 percent were currently unemployed.

@@

Testing the hypothesis necessitates defining what is considered a ``trivial effect size''.
Being a normative question, finding a clear, single, or unanimous answer is impossible.
However, it is still necessary and helpful to work toward a plausible benchmark.
I suggest the following SESOI for this research question:

@@

The hypothesis is analyzed using the interval testing approach as proposed by Dienes (2014).
To illustrate, if the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = -.02, {[}95\% CI: -.12, .08{]}), the hypothesis that the effect is trivial is supported.
If the confidence interval and the null region overlap (e.g., \emph{b} = -.22, {[}95\% CI: -.27, -.17{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence interval falls completely outside of the null-region (e.g., \emph{b} = -.40, {[}95\% CI: -.45, -.35{]}), the hypothesis is rejected and the existence of a meaningful positive effect is supported.
For an illustration, see Figure \ref{fig:sesoi}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/sesoi-1.pdf}
\caption{\label{fig:sesoi}Using confidence intervals to test a null region. In this study, a trivial effect of social media use on life satisfaction is defined as ranging from b = -.30 to b = .30}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa7}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{endnotes}
\setlength{\parskip}{0em}
\raggedbottom
\note{\clearpage}
\usepackage{setspace}
\AtBeginEnvironment{tabular}{\singlespacing}
\AtBeginEnvironment{lltable}{\singlespacing}
\AtBeginEnvironment{tablenotes}{\doublespacing}
\captionsetup[table]{font={stretch=1.5}}
\captionsetup[figure]{font={stretch=1.5}}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Effects of COVID-19 related social media use on well-being},
  pdflang={en-EN},
  pdfkeywords={COVID-19, well-being, social media, news use, panel study.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Effects of COVID-19 related social media use on well-being}
\author{Name blinded for review\textsuperscript{1}}
\date{}

\shorttitle{Effects of COVID-19 related social media use on well-being}

@@

\abstract{%
In times of crisis such as the COVID-19 pandemic, citizens need to stay informed about recent political events. To this end, people increasingly use social media. However, because social media are particularly engaging, many find it hard to disconnect, especially during times of crisis. Using data from the Austrian Corona Panel Project consisting of 3,485 participants from 34 waves, controlling for several stable and varying confounders, the results showed that COVID-19 related social media use did not meaningfully reduce well-being. Other factors such as health, income, exercise, or internal locus of control showed larger and meaningful effects.
}

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus?
In what region is it spreading?
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high (Statista, 2021).
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged (Sharma et al., 2022).
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children (Klein, 2021).
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health (Sandstrom et al., 2021).
Although research has begun addressing this question
(e.g., Bendau et al., 2021; Eden et al., 2020; Sewall et al., 2021),
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

This study investigates how different \emph{facets} of well-being are affected by different \emph{types} and different \emph{channels} of communication (Meier \& Reinecke, 2020).
Building on the typology of subjective well-being (Diener et al., 2018), three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used (Verduyn et al., 2015), I further distinguish three types of use and five popular channels.
The types of use include reading, liking and sharing, and posting COVID-19 related content.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting and liking COVID-19 related news.
Liking and sharing are combined as they both represent low-threshold, platform-ingrained, easily quantifiable interactions.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

\hypertarget{social-media-effects-on-well-being}{%
\subsection{Social Media Effects on Well-Being}\label{social-media-effects-on-well-being}}

How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time returning to prior levels (Sheldon \& Lucas, 2014).
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being (Lucas, 2007).
So although well-being can be affected by external events and factors, this does not happen easily.

Can social media use be such a factor?
Current literature overviews suggest that social media use on average does seem to decrease well-being (Meier \& Reinecke, 2020).
However, for most well-being outcomes, such as life satisfaction, general well-being, or loneliness, the effects are small (Meier \& Reinecke, 2020).
These small findings can be explained with the differential susceptibility of media effects model (Valkenburg \& Peter, 2013), which states that there is substantial \emph{variation} of media effects for individual users.
Whereas for some users social media are more beneficial, for others they are more harmful.
On average, however, and this is central for this study here, effects tend to be small (Valkenburg \& Peter, 2013).
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral (Beyens et al., 2021).
Whether or not effects are positive or negative depend on (a) dispositional factors (e.g., personality, temperament, gender), (b) developmental factors (e.g., age, developmental tasks), (c) and social factors (e.g., environment, norms, upbringing).
Finally, effects depend also on the content that is consumed.
If the content is aligned with dispositions, developmental capacities, and converging contexts, effects tend to be stronger (Valkenburg \& Peter, 2013).

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory (Zillmann, 1988), using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood (Marciano et al., 2022).
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.
On the other hand, although mood management theory suggests that effects should not be overly negative, it could be that although short-term effects are positive, long-term effects are negative.
Precisely because social media have so many positive consequences in the short run, this might cause problem in the long run---for example, because social media use displaces other potentially more meaningful and/or relaxing activities (Hall \& Liu, 2022).
As with many other things, there can be too much of a good thing.
It is therefore often asked whether social media can become addictive, and users sometimes express this fear themselves (Yang et al., 2021).
However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being (Duradoni et al., 2020).
In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media use represent a new regular behavior that should not be pathologized (Galer, 2018; van Rooij et al., 2018).

Second, while mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational (Katz et al., 1973).
Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.
If those beneficial media effects are missing, people will spend their time elsewhere.
And social media, in general, offer several beneficial effects, explaining why they are used that much.
They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself (Pelletier et al., 2020).
In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory this indirectly suggest that average effects on well-being are likely not particularly negative.

\hypertarget{social-media-during-covid-19}{%
\subsection{Social Media During COVID-19}\label{social-media-during-covid-19}}

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Access to information and resources:
Despite incorrect information, social media provide a vast platform for disseminating accurate and timely information about COVID-19 (John Hopkins University, 2023).
It allows individuals to stay informed about the latest updates, guidelines, and recommendations from reputable health organizations and experts.
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
Educational opportunities:
Social media can be a valuable educational tool, with various organizations and experts sharing informative content about COVID-19 (World Health Organization, 2023).
This can include explanatory videos, infographics, and articles that help individuals understand the virus, its transmission, prevention strategies, and vaccination information.
Access to educational resources on social media can empower individuals to make informed decisions regarding their health.
Community support and solidarity:
Social media platforms enable individuals to connect with others who are experiencing similar challenges during the pandemic (Guazzini et al., 2022).
Sharing experiences, offering support, and expressing solidarity can foster a sense of community and reduce feelings of isolation.
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
Promoting mental health and well-being:
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for maintaining mental well-being during the pandemic (Twitter, 2020).
They provide guidance on coping mechanisms, stress reduction techniques, and self-care practices.
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
Encouraging positive behaviors:
Social media campaigns and initiatives can promote positive behaviors related to COVID-19, such as mask-wearing, physical distancing, hand hygiene, and vaccination (Athey et al., 2023; Hunt et al., 2022).
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Misinformation and rumors:
Social media platforms can easily spread false or misleading information about COVID-19 (Li et al., 2020).
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
Overwhelm and anxiety:
Constant exposure to COVID-19-related content on social media can lead to information overload and contribute to heightened anxiety levels (J. Fan \& Smith, 2021).
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic (Sharma et al., 2022).
Confirmation bias and echo chambers:
Social media algorithms are designed to show users content that aligns with their interests and beliefs, potentially contributing to echo chambers (Cinelli et al., 2021).
This might lead to confirmation bias, where individuals are more likely to be exposed to information that confirms their preexisting beliefs.
In the context of COVID-19, this might reinforce false or misleading information and prevent individuals from considering alternative perspectives.
Cyberbullying and harassment:
Social media platforms are known for fostering negativity, with users sometimes engaging in cyberbullying and harassment (Giumetti \& Kowalski, 2022).
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
Social comparison and fear of missing out (FOMO):
Social media often showcase the highlights and accomplishments of others, encouraging social comparison (Przybylski et al., 2013).
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances (Sharma et al., 2022).
This might impact mental health, as most humans have an innate need for social support and physical presence, which online interactions might not fully replicate.

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison (Klein, 2021; Stainback et al., 2020; Yue et al., 2022).
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms (Bendau et al., 2021).
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes.
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being (Guazzini et al., 2022).
Additionally, engaging more on social media was associated with reduced feelings of loneliness (Latikka et al., 2022).
A couple of studies reported mostly neutral effects of social media use on well-being indicators (Bradley \& Howard, 2021; Eden et al., 2020; Sewall et al., 2021).
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being (see also Bradley \& Howard, 2021; Dörnemann et al., 2021; Liu \& Tong, 2020; Riehm et al., 2020; Sewall et al., 2021).

In conclusion, given these mixed empirical results, together with the observation that social media effects on well-being are very small in general, and that several plausible theoretical mechanisms exist for both positive and negative effects, I expect that COVID-19 related communication on social media should not be decidedly positive or negative.
It seems most likely that both positive and negative coexist, but that on average using social media for COVID-19 related reasons should not have substantial effects on well-being.

\begin{quote}
Hypothesis: The within-person effects of all measures of COVID-19 related social media use (types: reading, liking and sharing, posting; channels: Twitter, Instagram, Facebook, YouTube, WhatsApp) on all measures of well-being indicators (positive affect, negative affect, life satisfaction)---while controlling for several stable and varying covariates such as sociodemographic variables and psychological dispositions (see below)---will be trivial.
\end{quote}

@@

The data come from the Austrian Corona Panel Project (Kittel et al., 2021), which is a large-scale standalone panel study.
The data are hosted on AUSSDA, are publicly available here (\url{https://doi.org/10.11587/28KQNS}), and consist of 34 waves.
Participants were sampled from a pre-existing online access panel provided by the company Marketagent, Austria.
Panel members were incentivized with 180 credit points for each wave of the study.
The study was conducted between March 2020 and February 2023.
Between March 2020 and July 2020, the intervals between waves were weekly, until May 2022 (wave 32) monthly, and afterward after 5 months.
Each wave consists of at least 1,500 respondents.
Panel mortality was compensated through a continuous re-acquisition of new participants.
The sample size was \emph{N} = 3,641, with overall 123,794 observations.
For an overview of the study set-up, see Figure \ref{fig:study-desc}.

\begin{figure}
\includegraphics[width=1\textwidth]{figures/fig_study_description} \caption{Overview of study set-up}\label{fig:study-desc}
\end{figure}

Achieved via quota sampling, the sample matched the Austrian population in terms of age, gender, region/state, municipality size, and educational level.
In order to participate in the study, the respondents needed to be Austrian residents and had to be at least 14 years of age.
All respondents needed to have access to the internet (via computer or mobile devices such as smartphones or tablets).
Ethical review and approval was not required for the study in accordance with the local legislation and institutional requirements.
The participants provided their written informed consent to participate in this study.
The average age was 40 years, 49 percent were male, 14 percent had a University degree, and 5 percent were currently unemployed.

@@

Testing the hypothesis necessitates defining what is considered a ``trivial effect size''.
To this end, we need to define a so-called smallest effect size of interest (SESOI) (Lakens et al., 2018).
A trivial effect would then need to be smaller than the SESOI (see below).
What could be a minimally interesting, nontrivial effect?
Being a normative question, finding a clear, single, or unanimous answer is impossible.
However, it is still necessary and helpful to work toward a plausible benchmark.
I suggest the following SESOI for this research question:

@@

The hypothesis is analyzed using the interval testing approach as proposed by Dienes (2014).
To illustrate, let us consider the case of life satisfaction {[}SESOI: -.30 : +.30{]}.
If the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = -.05, {[}95\% CI: -.15, .05{]}), the hypothesis that the effect is trivial is supported.
If the confidence interval falls completely outside of the null-region (e.g., \emph{b} = -.40, {[}95\% CI: -.45, -.35{]}), the hypothesis is rejected and the existence of a meaningful negative effect is supported.
If the confidence interval and the null region overlap (e.g., \emph{b} = -.30, {[}95\% CI: -.35, -.25{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful positive effect is rejected.
If the confidence interval exceeds both sides of the null region (e.g., \emph{b} = -.025, {[}95\% CI: -.40, .35{]}), the hypothesis is not supported and judgement is suspended.
For an illustration, see Figure \ref{fig:sesoi}.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/sesoi-1.pdf}
\caption{\label{fig:sesoi}Using confidence intervals to test a null region. In this study, a trivial effect of social media use on life satisfaction is defined as ranging from b = -.30 to b = .30. Figure adapted from Dienes (2014).}
\end{figure}
